bad forecast good decision

### 2020-07-08 notebook/bad-forecast-good-decision.Rmd
reknit

```{r message=FALSE, warning=FALSE}
library(ggthemes)
library(hrbrthemes)
library(ggplot2)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
#theme_set(theme_solarized(base_size=16))

scale_colour_discrete <- function(...) scale_colour_solarized()
scale_fill_discrete <- function(...) scale_fill_solarized()
pal <- solarized_pal()(6)
txtcolor <- "#586e75"
```

```{r}
library(tidyverse)
library(MDPtoolbox)
library(expm)
library(mdplearning) # remotes::install_github("boettiger-lab/mdplearning")
```

```{r}
states <- seq(0,20, length.out = 200)
actions <- states
obs <- states
sigma_g <- 0.04
reward_fn <- function(x,h) pmin(x,h)
discount <- 0.99

# K is at twice max of f3; 8 * K_3 / 5
f1 <- function(x, h = 0, r = 2, K = 10 * 8 / 5){
  s <- pmax(x - h, 0)
  s + s * (r * (1 - s / K) )
}
f2 <- function(x, h = 0, r = 1, K = 10){
  s <- pmax(x - h, 0)
  s + s * (r * (1 - s / K) )
}

@@

f4  <- function(x, h = 0, r = 4e-7, K = 10){
  s <- pmax(x - h, 0)
  s + s ^ 8 * r * (1 - s / K) + s * (1 - s / K)
}

models <- list(f1 = f1, f2 = f2, f3 = f3)

true_model <- "f3"

## Could consider ricker-style versions of these models. need to take care to set limits correctly
## to get correspondence (recall x * e^y ~ x *(1 + y) for small y), e.g. something like
#f1 <- function(x, h = 0, r = .1, K = 15){
#  s <- pmax(x - h, 0)
#  s * exp(r * (1 - s / K) )
#}
#f3  <- function(x, h = 0, r = .01, K = 10){
#  s <- pmax(x - h, 0)
#  s * exp(s ^ 3 * r * (1 - s / K))
#}

@@

```{r}
d <- map_dfc(models, function(f) f(states) - states) %>% mutate(state = states)
d %>% pivot_longer(names(models), "model") %>%
  ggplot(aes(state, value, col=model)) +
  geom_point() + 
  geom_hline(aes(yintercept = 0)) + 
  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16))
```

```{r}
# A function to compute the transition matrices for each model:
transition_matrices <- function(f,
                      states,
                      actions,
                      sigma_g){

  transition <- array(0, dim = c(n_s, n_s, n_a))
  for (k in 1:n_s) {
    for (i in 1:n_a) {
      nextpop <- f(states[k], actions[i])
      if(nextpop <= 0){
        transition[k, , i] <- c(1, rep(0, n_s - 1))
      } else if(sigma_g > 0){
        x <- dlnorm(states, log(nextpop), sdlog = sigma_g)
        if(sum(x) == 0){ ## nextpop is computationally zero
          transition[k, , i] <- c(1, rep(0, n_s - 1))
        } else {
          x <- x / sum(x) # normalize evenly. 
          ## pile excess on boundary
          #N <- plnorm(states[n_s], log(nextpop), sigma_g)
          #x <- x * N / sum(x)
          #x[n_s] <- 1 - N + x[n_s]
          transition[k, , i] <- x
        }
      } else {
        stop("sigma_g not > 0")
      }
      reward[k, i] <- reward_fn(states[k], actions[i])
    }
  }
  transition
}

## Reward matrix is shared by each model
n_s <- length(states)
n_a <- length(actions)
reward <- array(0, dim = c(n_s, n_a))
for (k in 1:n_s) {
  for (i in 1:n_a) {
    reward[k, i] <- reward_fn(states[k], actions[i])
  }
}

transitions <- lapply(models, function(f) transition_matrices(f, states, actions, sigma_g))

@@

## Forecasting performance

@@

## Model analysis

policies <- map_dfr(transitions, function(P){
  soln <- mdp_value_iteration(P, reward, discount = discount)
  tibble(states, policy = soln$policy, escapement = states - actions[soln$policy])
}, .id = "model")

```{r}
policies %>%
  ggplot(aes(states,escapement, col=model, lty=model)) + geom_line(lwd=2)
```

Management under the wrong model (`f3`)

```{r}
library(mdplearning)
Tmax <- 100
x0 <- which.min(abs(states - 6))
reps <- 100
set.seed(12345)

```{r}
fig_ts <- 
  sims %>%
  filter(time < 25, reps < 5) %>%
  ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 
  geom_line(alpha=0.5) 
fig_ts
```

Note that the pattern is not driven merely be differences in the unfished equilibrium.  Compare forecast performance under management.  Score is the sum of the log probabilities of observed outcomes, given the

```{r}
scores <- sims %>%
  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
  filter(model < 3) %>%
  select(time, state_index, next_state, action_index, model, reps) %>% 
  rowwise() %>%
  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
  group_by(reps, model) %>%
  summarise(score = sum(log(prob)))

```{r fig.cap = "Corresponding utility (measured as mean net present value, that is: cumulative value, discounting future values by $\\delta^t$, averaged across replicates). Note that the exepcted utility under model 1, which has the worst forecast, is nearly identical to the optimal utility achieved by managing under the correct model, 3.  The utility derived from model 2 is far smaller, despite it's overall better performance in long term forecasts."}
##  Net Present Value accumulates over time, equivalent for models with near-identical management stategy
npv_df <- sims %>% 
  group_by(model, reps) %>%
  mutate(npv = cumsum(value * discount ^ time)) %>%
  group_by(time, model)  %>% 
  summarise(mean_npv = mean(npv))
npv_df %>%
  ggplot(aes(time, mean_npv, fill=model)) +
  ## Area plots are weird but avoids overplotting
  geom_area(alpha = 0.9) +
   #geom_ribbon(aes(ymin = 0, ymax = mean_npv), alpha = 0.6) + 
  ylab("Net present value") + xlab("time")

Model with the egregiously optimistic long-term forecast for the stock size of an unexploited fishery nevertheless actually leads to more conservative management.  In contrast, the model which correctly predicts the long-term average stock size without fishing nevertheless leads to substantial overharvesting.

>>>>

```{r graphics_setup, message=FALSE, warning=FALSE}
## Plotting theme settings -- aesthetics only, all can be omitted without consequence
library(ggthemes)
library(hrbrthemes)
library(ggplot2)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
#theme_set(theme_solarized(base_size=16))

scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
pal <- ggthemes::solarized_pal()(6)
txtcolor <- "#586e75"
```

```{r setup, message=FALSE}
## Required computational libraries
library(tidyverse)
library(MDPtoolbox)
library(expm)
library(mdplearning) # remotes::install_github("boettiger-lab/mdplearning")
```

```{r model_definitions}
states <- seq(0,24, length.out = 240)
actions <- states
obs <- states
sigma_g <- 0.05
reward_fn <- function(x,h) pmin(x,h)
discount <- 0.99

# K is at twice max of f3; 8 * K_3 / 5
f1 <- function(x, h = 0, r = 2, K = 10 * 8 / 5){
  s <- pmax(x - h, 0)
  s + s * (r * (1 - s / K) )
}
f2 <- function(x, h = 0, r = 0.5, K = 10){
  s <- pmax(x - h, 0)
  s + s * (r * (1 - s / K) )
}

@@

# looks even closer to f2; not used
#f4  <- function(x, h = 0, r = 4e-7, K = 10){
#  s <- pmax(x - h, 0)
#  s + s ^ 8 * r * (1 - s / K) + s * (1 - s / K)
#}

models <- list(f1 = f1, f2 = f2, f3 = f3)
model_sigmas <- c(sigma_g, 1.5 * sigma_g, sigma_g)

true_model <- "f3"
```

```{r transition_matrices}
# A function to compute the transition matrices for each model:
transition_matrices <- function(f,
                      states,
                      actions,
                      sigma_g){

  transition <- array(0, dim = c(n_s, n_s, n_a))
  for (k in 1:n_s) {
    for (i in 1:n_a) {
      nextpop <- f(states[k], actions[i])
      if(nextpop <= 0){
        transition[k, , i] <- c(1, rep(0, n_s - 1))
      } else if(sigma_g > 0){
        x <- dlnorm(states, log(nextpop), sdlog = sigma_g)
        if(sum(x) == 0){ ## nextpop is computationally zero
          transition[k, , i] <- c(1, rep(0, n_s - 1))
        } else {
          x <- x / sum(x) # normalize evenly
          transition[k, , i] <- x
        }
      } else {
        stop("sigma_g not > 0")
      }
      reward[k, i] <- reward_fn(states[k], actions[i])
    }
  }
  transition
}

## Compute reward matrix (shared across all models)
n_s <- length(states)
n_a <- length(actions)
reward <- array(0, dim = c(n_s, n_a))
for (k in 1:n_s) {
  for (i in 1:n_a) {
    reward[k, i] <- reward_fn(states[k], actions[i])
  }
}

## Now compute the transition matrices for each model
transitions <- lapply(seq_along(models), 
                      function(i) transition_matrices(models[[i]], 
                                                      states, 
                                                      actions, 
                                                      model_sigmas[[i]]))
names(transitions) <- c("f1", "f2", "f3")

@@

```{r sdp, results="hide"}
## This is the most (only) computationally expensive code chunk:
policies <- 
  map_dfr(transitions, 
          function(P){
    soln <- mdp_value_iteration(P, reward, discount = discount)
    escapement <- states - actions[soln$policy]
    tibble(states, policy = soln$policy, escapement)
    }, 
  .id = "model")
```

```{r simulations}
library(mdplearning)
Tmax <- 100
x0 <- which.min(abs(states - 6))
reps <- 100
set.seed(12345)

@@

## Forecast skill

Note the low variability in the predicted value is a result of the decision strategy, which tweaks harvest up or down with variation.

```{r plot_stepahead}
compare <- sims %>%
  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
  rowwise() %>%
  mutate(prob =  transitions[[model]][state_index, next_state, action_index],
         expected = transitions[[model]][state_index, , action_index]  %*% states,
         low = states[ max(which(cumsum(transitions[[model]][state_index,,action_index]) < 0.025)) ],
         high = states[ min(which(cumsum(transitions[[model]][state_index,,action_index]) > 0.975)) ],
         true = states[ next_state],
         model = as.character(model)) %>%
 select(time, model, true, expected, low, high, prob, reps)

compare %>% filter(model != "3") %>%  filter(reps == 3, time < 10) %>%
  ggplot(aes(time, col = model, fill = model)) + 
  geom_point(aes(y = true), pch = "*", size = 12) +
  geom_point(aes(y = expected)) + 
  geom_errorbar(aes(ymin = low, ymax = high))
```

```{r plot_stepahead_unfished}
compare_unfished <- sims
compare_unfished$state_index <- rep(sims$state_index[sims$model == "1"],3)
compare_unfished <- compare_unfished %>% 
  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
  rowwise() %>%
  mutate(expected = transitions[[model]][state_index, , 1]  %*% states,
         low = states[ max(which(cumsum(transitions[[model]][state_index,,1]) < 0.025)) ],
         high = states[ min(which(cumsum(transitions[[model]][state_index,,1]) > 0.975)) ],
         true = states[ next_state ],
         model = as.character(model)) %>% filter(reps == 2, time < 10, model != "3")

compare_unfished %>%
  ggplot(aes(time, col = model, fill = model)) + 
  geom_point(aes(y = expected)) + 
  geom_errorbar(aes(ymin = low, ymax = high)) +
  geom_point(aes(y = true), pch = "*", size = 12, col = "grey20", show.legend = FALSE)
```

```{r prob_scores, message = FALSE}
scores <- sims %>%
  mutate(next_state = dplyr::lead(state_index), model = as.integer(model)) %>%
  filter(model < 3) %>%
  rowwise() %>%
  mutate(prob =  transitions[[model]][state_index, next_state, action_index] ) %>% 
  group_by(reps, model) %>%
  summarise(score = sum(log(prob)))

## Simulations

```{r plot_sims}
fig_ts <- 
  sims %>%
  filter(time < 25, reps < 5) %>%
  ggplot(aes(time, state, col=model, group = interaction(model,reps))) + 
  geom_line(alpha=0.3) 
fig_ts
```

```{r plot_npv, message = FALSE, fig.cap = "Corresponding utility (measured as mean net present value, that is: cumulative value, discounting future values by $\\delta^t$, averaged across replicates). Note that the exepcted utility under model 1, which has the worst forecast, is nearly identical to the optimal utility achieved by managing under the correct model, 3.  The utility derived from model 2 is far smaller, despite it's overall better performance in long term forecasts."}
##  Net Present Value accumulates over time, equivalent for models with near-identical management stategy
npv_df <- sims %>% 
  group_by(model, reps) %>%
  mutate(npv = cumsum(value * discount ^ time)) %>%
  group_by(time, model)  %>% 
  summarise(mean_npv = mean(npv)) %>% 
  arrange(model, time) %>% ungroup()

optimal <- select(filter(npv_df, model == "3"), time, mean_npv)

npv_df %>%
  filter(model != "3", time %in% seq(1,100, by = 5)) %>%
  ggplot(aes(time, mean_npv)) +
  geom_line(data = optimal, lwd = 1.5, col = "grey20") +
  geom_point(aes(col=model), size = 4, alpha = 0.8) + 
  ylab("Net present value") + xlab("time")

```{r plot_models}
d <- map_dfc(models, function(f) f(states) - states) %>% mutate(state = states)
d %>% pivot_longer(names(models), "model") %>%
  ggplot(aes(state, value, col=model)) +
  geom_point() + 
  geom_hline(aes(yintercept = 0)) + 
  coord_cartesian(ylim = c(-5, 8), xlim = c(0,16))
```

```{r plot_policies}
policies %>%
  ggplot(aes(states,escapement, col=model, lty=model)) + geom_line(lwd=2)
```

Model with the egregiously optimistic long-term forecast for the stock size of an unexploited fishery nevertheless actually leads to more conservative management.  In contrast, the model which correctly predicts the long-term average stock size without fishing nevertheless leads to substantial over-harvesting.

# Appendix

### Unharvested forecasting performance

Here we examine the forecasts under each model, `Tmax` steps into the future, 
ignoring harvesting.  This largely highlights only the differences in the 
unfished equilibria of the models.  It less surprising that bad performance
in predicting the long-term equilibrium would not necessarily mean bad management.

@@
